---
title: "Gene expression data of human hepatocellular carcinoma (HCC)"
author: "Tirdod Behbehani, Elisa Scocco,Iñigo Exposito"
date: "2025-04-03"
output: html_document
experiment type: Expression profiling by array
Link: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE14520
---

```{r}
# Install Bioconductor
install.packages("BiocManager")
BiocManager::install("GEOquery")
BiocManager::install("biomaRt")
```

```{r}
# Load libraries
library(GEOquery)
libraray(biomaRt)
```


#################### GEN EXPRESSION ##################################

1. Download the gen expression data and export to CSV

```{r}
# Download the dataset
gse <- getGEO("GSE14520", GSEMatrix = TRUE)

# Check the downloaded data
exprs_data <- exprs(gse[[1]])  # Get the expression data

#write csv
write.csv(exprs_data, "GSE14520_expression_data.csv")
```

2. Read the CSV

```{r}
#Read CSV
gen_expression<-read.csv('/Users/inigo/GSE14520_expression_data.csv',row.names=NULL)
```

The first column contains the gen identifier. The other columns represent a sample (it can be biological individual). The values in these columns show how much that gene is expressed in each sample

```{r}
#Check results
head(gen_expression)
```
Get the transposed matrix
```{r}
# Remove the index colum and keep the gene names in the first column
gen_expression_cleaned <- gen_expression[, -1]

# Set the row names to be the gene names 
rownames(gen_expression_cleaned) <- gen_expression$X

# Now transpose the data 
gen_data <- t(gen_expression_cleaned)

# Convert the transposed data to a data frame
gen_data <- as.data.frame(gen_data)

# Add an ID column to the transposed data
gen_data$ID <- rownames(gen_data)

# Move the 'ID' column to the first position
gen_data <- gen_data[, c(ncol(gen_data), 1:(ncol(gen_data) - 1))]

# Check the result
head(gen_data)

```

################ CLINICAL DATA #######################################

```{r}
# Define the path to the TXT file
file_path <- "/Users/inigo/Downloads/GSE14520_Extra_Supplement.txt"

# Read the file into R
clinical_data <- read.table(file_path, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```
 
Here we also have the GSM, which is the same as in the previous data set.
```{r}
head(clinical_data)
```

```{r}
#Check names of columns and types of covariates
colnames(clinical_data)
str(clinical_data)
```
#Categorical variables

```{r}
unique(clinical_data[4]) #Tissue.Type  
unique(clinical_data[5]) #Prediction of the risk of metastasis 
unique(clinical_data[7]) #Survival groups
unique(clinical_data[8]) #Gender
unique(clinical_data[10]) #Status of hepatitis B 
unique(clinical_data[11]) #Alanine Aminotransferase (enzime)
unique(clinical_data[12]) #Main tumor size
unique(clinical_data[13]) #multinodular (formation of multiple nodules in the liver)
unique(clinical_data[14]) #Cihrrosis (chronic liver disease)
unique(clinical_data[15]) #Stage of liver cancer
unique(clinical_data[16]) #Stage of liver cancer
unique(clinical_data[17]) #Stage of liver cancer
unique(clinical_data[18]) #Alpha-Fetoprotein (AFP) level 
unique(clinical_data[19]) #Survival status
unique(clinical_data[21]) #Recurrence status (whether the disease has come back after treatment)

```

########## DATA PREPARATION ######################################

```{r}
# Merge the transposed expression data with clinical data on the Sample_ID column
merged_data <- merge(gen_data, clinical_data, by.x = "ID", by.y = "Affy_GSM", all.x = TRUE)
# View the first few rows of the merged data
head(merged_data)
```
```{r}
# Connect to Ensembl Biomart
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

```

We will change the gen expression by using the RNA_seq instead og the microarray
```{r}
# Identify columns that end with "at"
at_columns <- grep("at$", colnames(merged_data), value = TRUE)

# Print the "at" columns to check
print(at_columns)

```


```{r}
#Get the same format for gen ID
gene_info_ens <- getBM(attributes = c("affy_hg_u133_plus_2", "ensembl_gene_id"),
                       filters = "affy_hg_u133_plus_2", 
                       values = at_columns, 
                       mart = ensembl)

```

Join all the information

```{r}
#Check for empty genes
empty_genes <- gene_info_ens[, 2] == ""  # Empty strings check
count_missing <- sum(empty_genes)
valid_mapping <- gene_info_ens[!empty_genes, ]
# Create a mapping between the microarray gene ID and Ensembl gene ID
id_mapping <- setNames(valid_mapping$ensembl_gene_id, valid_mapping$affy_hg_u133_plus_2)

# Replace the column names in merged_data
colnames(merged_data)[colnames(merged_data) %in% names(id_mapping)] <- id_mapping[colnames(merged_data)[colnames(merged_data) %in% names(id_mapping)]]

# Verify the updated column names
head(merged_data)
```

```{r}
# Export the entire merged_data dataframe to a CSV file
write.csv(merged_data, "liver_cancer.csv", row.names = FALSE)
```






